Understanding meta-analysis through data simulation with applications to power analysis

### 2023-07-03 documents/paper/paper.Rmd
adding equal-effects and fix equations

A meta-analysis is an essential tool for combining knowledge from multiple studies quantitatively. Meta-analysis is commonly used together with a systematic review of the literature. The meta-analysis has several advantages. Firstly, it allows combining evidence from multiple studies assigning more weight to studies with lower estimation variability. Then using meta-regression, it is possible to assess the impact of study-level variables that could impact the phenomenon. Finally, considering the replication crisis, there are statistical methods to determine the presence and extent of the publication bias. Despite the advantages, meta-analysis implementation is not always straightforward, especially for complex data structures. Additionally, the fact that there are several introductory and advanced resources to understand meta-analysis [@Borenstein2009-mo; @Schmid2022-xe; @Harrer2021-bz], to our knowledge, there are no introductory resources about how to simulate realistic meta-analytic data. Simulating data has several advantages because it requires understanding the statistical method and the data-generation process. Furthermore, data simulation is the primary tool when it comes to evaluating a new analysis method, estimating the statistical power, or understanding the long-run behavior of our data generation process [@Ingalls2011-xt; @Gelman2020-tg, pp. 69-76; @Gelman2006-pc, pp. 155-176]. A recent paper by DeBruine and Barr [-@DeBruine2021-id], which deeply inspired the current work, proposed a stimulating way to understand linear mixed effects via data simulation. Simulating data is also a powerful educational tool within this framework. For these reasons, this work aims to introduce the basic concepts of meta-analysis and Monte Carlo simulations for fixed and random-effects, and meta-regression models with applications also for statistical power calculation. In the first section, we will introduce basic concepts of the meta-analysis that are useful for setting up the simulation. We will evaluate the effect size, variance calculation, and the fixed vs. random effects model distinction. Then we will describe how to simulate data for these models and simulate a meta-regression model with a categorical and numerical predictor. Finally, we will introduce the power analysis extending the previous examples to estimate the statistical power. We used the R statistical programming language `r rcit`. We assume the reader is familiar with the basic concepts of R, but core functions will be explained. Code and materials are available on the OSF repository ([https://osf.io/54djn/](https://osf.io/54djn/)). Multivariate and multilevel models’ simulation examples and more details about the coding approach are available in the supplementary materials.

@@

### Fixed vs random-effects model

The core of a meta-analysis is combining the results of multiple studies giving more weight to studies that provide a more precise effect estimation. We can divide meta-analysis models into two categories: the *fixed-effects* and the *random-effects model*. If we imagine combining multiple studies replicating the same experiment, we assume the real treatment effect $\theta$ is *fixed,* and each study is a more or less precise estimation of this parameter. This is an essential and often inappropriate assumption because it ignores that study-level characteristics may impact the actual treatment effect.

Suppose some study-level characteristics (e.g., participants' age, sex, or socioeconomic status) or the experimental paradigm (e.g., type of memory task or difficulty) could impact the treatment effect. In that case, the assumption of the *fixed-effects model* is no longer plausible. There is no unique real effect but a distribution of effects with mean $\mu_{\theta}$ and variance $\tau^{2}$. The mean is the *average* effect, and the variance is the heterogeneity of the phenomenon. The *random-effects model* will estimate the mean ($\mu_{\theta}$) and the heterogeneity $\tau^{2}$ will help generalize to the population of effects beyond the pool of included studies.

Most importantly, the parameters estimated with the *fixed* and *random-effects* do not represent the same quantity. The *random-effects model* estimates the *average* effect across a distribution of true effects, while the *fixed-effects model* estimates a single true effect. Figure 1 depicts the idea of the fixed vs. random-effects model. The *fixed-effects model* can be considered a particular type of *random-effects model* where $\tau^{2} = 0$ thus, the average effect is also the unique true effect.

@@

We load relevant packages and set the `seed` for reproducibility to prepare the R environment for the simulations.

```{r sim-setup, eval = FALSE}
library(dplyr) # for data manipulation
library(tidyr) # for data manipulation
library(metafor) # for meta-analysis models fitting
seed <- 2023 # general seed for all simulations
```

@@

\begin{align}
\begin{gathered}
y_i = \overline{T} - \overline{C}
(\#eq:effsize)
\end{gathered}
\end{align}

\begin{align}
\begin{gathered}
\sigma_{\epsilon_i}^{2} = \frac{s^2_T}{n_T} + \frac{s^2_C}{n_C}
(\#eq:effsizevar)
\end{gathered}
\end{align}

@@

```{r notation-tab, echo=FALSE, result = "asis"}
notation_table <- tibble::tribble(
     ~`Equation`,     ~`Code`,                                               ~`Description`,
       "$\\theta$",   "theta",                  "Fixed-effects model real effect size",
       "$\\mu$",      "mu",                        "Random-effects model real effect size",
        "$\\tau^2$",    "tau2",                            "Effect size heterogeneity",
       "$\\tau^2_r$",   "tau2r",                   "Residual effect size heterogeneity",
  "$\\bar y_{T, C}$",        "",               "Mean of the experimental/control group",
      "$s_{T, C}$",        "", "Standard deviation of the experimental/control group",
        "$n_{T, C}$",  "nt, nc",        "Sample size of the experimental/control group",
             "$d_i$",      "yi",                                 "Observed effect size",
      "$\\sigma_{\\epsilon_i}^{2}$",      "vi",                          "Observed sampling variance",
       "$\\theta_i$", "theta\\_i",              "Random effect for the study i",
        "$\\beta_0$",      "b0",                            "Meta-regression intercept",
         "$\\beta_1$",      "b1",                                "Meta-regression slope",
               "$k$",       "k",                                    "Number of studies",
     "$\\epsilon_i$",        "",             "Sampling error for the study i"
  )

@@

Now we can fit the meta-regression model with the `rma` function as for the random-effects model with the addition of `mods = ~ exp` that indicates which variable/s to consider as moderators. The results are presented in Table \@ref(tab:res-meta-reg-dummy). Now the model will estimate an *intercept* parameter (i.e., $\beta_0$) that is the value of $y$ when $X_1$ is zero (i.e., for lab-based studies) or in other terms the expected value for lab-based studies. Then the $\beta_1$ parameter represents the estimated difference in $y$ between the values of $X_1$ (i.e., lab-based vs online experiments). As said before, $\tau^2$ is now the residual heterogeneity that is interpreted as the variability between effect sizes after controlling for the moderator $X_1$.

Now we can fit the meta-regression model with the `rma` function as for the random-effects model with the addition of mods = `~ exp` that indicates which variable/s to consider as moderators. The results are presented in Table \@ref(tab:res-meta-reg-dummy). Next, the model will estimate an *intercept* parameter (i.e., $\beta_{0}$) that is the value of $y$ when $X_{1}$ is zero (i.e., for lab-based studies) or, in other terms, the expected value for lab-based studies. Then the $\beta_{1}$ parameter represents the estimated difference in $y$ between the values of $X_{1}$ (i.e., lab-based vs. online experiments). As mentioned above, $\tau^{2}$ the residual heterogeneity is now interpreted as the variability between effect sizes after controlling for the moderator $X_{1}$.

@@

(ref:img-fixed-vs-random) The difference between the assumptions of the *fixed* and *random-effects model*. Each distribution depicts the sampling distribution of $k = 4$ hypothetical studies ($i = 1$, $2$, ..., $4$) with a certain observed effect size $y_{i}$ (red squares), sampling variability $\sigma_{i}^{2}$ and the 95% confidence interval (black segment). The *fixed-effects* plot on the left suggests that each observed effect size has the same underlying true effect $\theta$ (each distribution has the same mean) with a different degree of precision (e.g., Study 1 is more precise than Study 3). In practice, each study has a different observed effect size where studies with high precision (i.e., narrow sampling distributions) will be remarkably close to the real effect ($\theta$). The *random-effects model* on the right suggests that beyond the error term ($\epsilon_{i}$), each real effect size is composed of a fixed part (now $\mu_{\theta}$) and a random part ($\theta_{i}$, black segments below the sampling distributions). The random part is determined by $\tau^{2}$, which represents the variability of the effect size distribution at the population level. When $\tau^{2}$ is zero, the random-effects reduces to a fixed-effects model.

>>>>

A meta-analysis is an essential tool for combining knowledge from multiple studies quantitatively. Meta-analysis is commonly used together with a systematic review of the literature. The meta-analysis has several advantages. Firstly, it allows combining evidence from multiple studies assigning more weight to studies with lower estimation variability. Then using meta-regression, it is possible to include variables (i.e., moderators) to explain the observed heterogeneity [@Borenstein2009-mo, pp.187-203]. More recently, location-scale models have been developed to inlcude predictors also on the residual heterogeneity [@Viechtbauer2022-rz]. Finally, considering the replication crisis, there are statistical methods to determine the presence and extent of the publication bias. Despite the advantages, meta-analysis implementation is not always straightforward, especially for complex data structures. Additionally, the fact that there are several introductory and advanced resources to understand meta-analysis [@Borenstein2009-mo; @Schmid2022-xe; @Harrer2021-bz], to our knowledge, there are no introductory resources about how to simulate realistic meta-analytic data. Simulating data has several advantages because it requires understanding the statistical method and the data-generation process. Furthermore, data simulation is the primary tool when it comes to evaluating a new analysis method, estimating the statistical power, or understanding the long-run behavior of our data generation process [@Ingalls2011-xt; @Gelman2020-tg, pp. 69-76; @Gelman2006-pc, pp. 155-176]. A recent paper by DeBruine and Barr [-@DeBruine2021-id], which deeply inspired the current work, proposed a stimulating way to understand linear mixed effects via data simulation. Simulating data is also a powerful educational tool within this framework. For these reasons, this work aims to introduce the basic concepts of meta-analysis and Monte Carlo simulations for fixed and random-effects, and meta-regression models with applications also for statistical power calculation. In the first section, we will introduce basic concepts of the meta-analysis that are useful for setting up the simulation. We will evaluate the effect size, variance calculation, and the fixed vs. random effects model distinction. Then we will describe how to simulate data for these models and simulate a meta-regression model with a categorical and numerical predictor. Finally, we will introduce the power analysis extending the previous examples to estimate the statistical power. We used the R statistical programming language `r rcit`. We assume the reader is familiar with the basic concepts of R, but core functions will be explained. Code and materials are available on the OSF repository ([https://osf.io/54djn/](https://osf.io/54djn/)). Multivariate and multilevel models’ simulation examples and more details about the coding approach are available in the supplementary materials.

@@

### Equal vs fixed vs random-effects model

The core of a meta-analysis is combining the results of multiple studies giving more weight to studies that provide a more precise effect estimation. We can find essentially three meta-analysis models: the equal-effects, the fixed-effects and the random-effects model [@Laird1990-ro, @Hedges1998-uo]. The *equal-effects* model assume that each study included in the meta-analysis is a more or less precise estimation of the true underlying effect ($\theta$). In other terms, we are assuming that there is no variability (i.e., heterogeneity) among true effect sizes. On the other side, suppose some study-level characteristics (e.g., participants' age, sex, or socioeconomic status) or the experimental paradigm (e.g., type of memory task or difficulty) could impact the treatment effect. The *random-effects* model assume a distribution of real effects with mean $\mu_{\theta}$ and variance $\tau^{2}$. Finally, the *fixed-effects* estimate the average effect of the included pool of studies ignoring the presence of heterogeneity. While the distinction between the equal and fixed-effects model is theoretical, the estimated model is the exactly the same^[see https://wviechtb.github.io/metafor/reference/misc-models.html for a detailed explanation]. The *random-effects* model assume and estimate heterogeneity among effect sizes leading to different model parameters. From an inferential point of view, the *random-effects* model provide unconditional inference to the population of effect sizes while the *fixed-effects* model estimate the average effect of the selected studies, and not population-level parameters. The *equal-effects* model assume a single true population effect to be estimated. In this tutorial we will consider the equal and random-effects models.

@@

This tutorial use several R packages for the simulations, meta-analysis fitting and figures/tables. For the data manipulation we used the *tidyverse* [@R-tidyverse] package. For the models fitting we used the *metafor* [@metafor2010] package. For figures and tables the *ggplot2* [@ggplot22016] or the `metafor::forest()` function, *kableExtra* [@R-kableExtra] and *papaja* [@R-papaja] packages. We set the `seed` for reproducibility to prepare the R environment for the simulations.

```{r sim-setup, eval = FALSE}
library(dplyr) # for data manipulation (within tidyverse)
library(tidyr) # for data manipulation (within tidyverse)
library(metafor) # for meta-analysis models fitting
seed <- 2023 # general seed for all simulations
```

@@

\begin{align}
\begin{gathered}
D = \overline{T} - \overline{C}
(\#eq:effsize)
\end{gathered}
\end{align}

\begin{align}
\begin{gathered}
\sigma_{\epsilon D}^{2} = \frac{s^2_T}{n_T} + \frac{s^2_C}{n_C}
(\#eq:effsizevar)
\end{gathered}
\end{align}

@@

```{r notation-tab, echo=FALSE, result = "asis"}
notation_table <- tibble::tribble(
     ~`Equation`,     ~`Code`,                                               ~`Description`,
       "$\\theta$",   "theta",                  "Fixed-effects model real effect size",
       "$\\mu$",      "mu",                        "Random-effects model real effect size",
        "$\\tau^2$",    "tau2",                            "Effect size heterogeneity",
       "$\\tau^2_r$",   "tau2r",                   "Residual effect size heterogeneity",
  "$\\overline{T}, \\overline{C}$",        "",               "Mean of the experimental/control group",
      "$s_{T, C}$",        "", "Standard deviation of the experimental/control group",
        "$n_{T, C}$",  "nt, nc",        "Sample size of the experimental/control group",
             "$d_i$",      "yi",                                 "Observed effect size",
      "$\\sigma_{\\epsilon_i}^{2}$",      "vi",                          "Observed sampling variance",
       "$\\theta_i$", "theta\\_i",              "Random effect for the study i",
        "$\\beta_0$",      "b0",                            "Meta-regression intercept",
         "$\\beta_1$",      "b1",                                "Meta-regression slope",
               "$k$",       "k",                                    "Number of studies",
     "$\\epsilon_i$",        "",             "Sampling error for the study i"
  )

@@

Now we can fit the meta-regression model with the `rma` function as for the random-effects model with the addition of `mods = ~ exp` that indicates which variable/s to consider as moderators. The results are presented in Table \@ref(tab:res-meta-reg-dummy). Now the model will estimate an *intercept* parameter (i.e., $\beta_0$) that is the value of $y$ when $X_1$ is zero (i.e., for lab-based studies) or in other terms the expected value for lab-based studies. Then the $\beta_1$ parameter represents the estimated difference in $y$ between the values of $X_1$ (i.e., lab-based vs online experiments). As said before, $\tau^2_r$ is now the residual heterogeneity that is interpreted as the variability between effect sizes after controlling for the moderator $X_1$.

Now we can fit the meta-regression model with the `rma` function as for the random-effects model with the addition of mods = `~ exp` that indicates which variable/s to consider as moderators. The results are presented in Table \@ref(tab:res-meta-reg-dummy). Next, the model will estimate an *intercept* parameter (i.e., $\beta_{0}$) that is the value of $y$ when $X_{1}$ is zero (i.e., for lab-based studies) or, in other terms, the expected value for lab-based studies. Then the $\beta_{1}$ parameter represents the estimated difference in $y$ between the values of $X_{1}$ (i.e., lab-based vs. online experiments). As mentioned above, $\tau^{2}_r$ the residual heterogeneity is now interpreted as the variability between effect sizes after controlling for the moderator $X_{1}$.

@@

(ref:img-fixed-vs-random) The difference between the assumptions of the *fixed* and *random-effects model*. Each distribution depicts the sampling distribution of $k = 4$ hypothetical studies ($i = 1$, $2$, ..., $4$) with a certain observed effect size $y_{i}$ (red squares), sampling variability $\sigma_{i}^{2}$ and the 95% confidence interval (black segment). The *fixed-effects* plot on the left suggests that each observed effect size has the same underlying true effect $\theta$ (each distribution has the same mean) with a different degree of precision (e.g., Study 1 is more precise than Study 3). In practice, each study has a different observed effect size where studies with high precision (i.e., narrow sampling distributions) will be remarkably close to the real effect ($\theta$). The *random-effects model* on the right suggests that beyond the error term ($\epsilon_{i}$), each real effect size is composed of a fixed part (now $\mu_{\theta}$) and a random part ($\theta_{i}$) sampled from a normal distribution with mean $\mu_{\theta}$  and variance $\tau^{2}$. When $\tau^{2}$ is zero, the random-effects reduces to a fixed-effects model.
